% Compile with this shell command:
% latex dougo-resume && dvipdfm dougo-resume && tex2page dougo-resume

\documentclass{article}
\usepackage{times}
\usepackage{tex2page}

\setlength{\textheight}{9in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\footskip}{0in}

\setlength{\textwidth}{6.5in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}

\pagestyle{empty}

\begin{document}

\def\TZPtitle{Doug Orleans}

\htmlonly
{\small [This document is also available as \urlh{dougo-resume.pdf}{PDF},
\urlh{dougo-resume.dvi}{DVI}, and \urlh{dougo-resume.tex}{\LaTeX}.]}
\endhtmlonly

\begin{center}
{\urlh{http://steak.place.org/dougo/}{\LARGE \textbf{Doug Orleans}}} \\
{\large 882 Broadway \#3R \\ Somerville, MA 02144} \\
dougorleans@gmail.com \\
(978) 987-5084
\texonly \small
\\ http://steak.place.org/dougo/resume/
\\ http://www.linkedin.com/in/dougo
\\ http://www.ohloh.net/accounts/dougo
\endtexonly
\htmlonly
\\ \urlh{http://www.linkedin.com/in/dougo}{LinkedIn}
\urlh{http://www.ohloh.net/accounts/dougo}{Ohloh}
\endhtmlonly
\end{center}

\section*{Employment}

\begin{description}
\item[Dec. 2011--present]
{\large \urlh{http://mcna.net/}{MCNA Dental}}, Fort Lauderdale, FL \\
Developed and maintained various features of DentalTrac, a Ruby on
Rails web application used by the entire company for managing dental
insurance claims, subscribers, and providers.  Projects worked on:
\begin{itemize}
  \item Building a Rails front-end for a Lisp/PostgreSQL-based
    workflow/business-process management system.
  \item Building a case management tool for the fraud, waste, and
    abuse department.
  \item Complying with state regulations for new dental plans (Texas
    CHIP and Medicaid).
  \item Migrating the database from MySQL to PostgreSQL and updating
    db-specific dependencies in the code.
  \item Upgrading the app to use Rails 3.1.1 (from 0.12) and modern
    Rails coding practices.
  \item Generally improving reliability, usability, and performance.
\end{itemize}

\item[Jun. 2009--Dec. 2011]
{\large \urlh{http://appleseed-sc.com/}{Appleseed Software Consulting}},
Somerville, MA \\
As an offsite contract consultant, implemented web applications for
clients using the Catalyst framework in Perl, along with MySQL,
Template Toolkit, FormFu, and jQuery.
\begin{itemize}
\item For \urlh{http://dca.tufts.edu/}{Tufts University Digital
  Collections \& Archives}, helped design and implement
  \urlh{https://github.com/dcatufts/sabt}{TAPER}, a web tool for
  managing XML accession requests, and CIDER, a web tool for managing
  hierarchical collections.
\item For \urlh{http://www.mainehistory.org/}{Maine Historical
  Society}, helped implement various pages on their public-facing
  website for displaying and managing databases of archival records.
\item For various other clients, helped implement web-based
  applications for managing databases.
\end{itemize}

\item[Jul. 2010--Sep. 2010]
{\large \urlh{http://aceditions.com/}{Art+Culture Editions}},
New York, NY \\
As an offsite contract consultant, helped implement an e-commerce web
application with Ruby on Rails and Spree.

\item[Jan. 2005--Nov. 2007]
{\large Gensym Corporation}, Burlington, MA \\
Member of the \urlh{http://www.gensym.com/?p=what_it_is_g2}{G2}
Core development team.  Maintained and added features to a
20-year-old codebase of over one million lines of Common Lisp.
Designed and implemented client and server libraries for HTTP, SOAP,
WSDL.  Helped implement the G2 Graphical Language, based on WS-BPEL.

\item[1996--2003] 
{\large Research assistant, Northeastern University}, Boston, MA \\
Member of Dr. Karl Lieberherr's
\urlh{http://www.ccs.neu.edu/research/demeter/}{Demeter} research
team.  Designed, implemented, and maintained DemeterJ, an extension to
the Java language for adaptive programming (an early form of
aspect-oriented programming).  Later, took over development of
\urlh{../papers/dj/}{DJ} (Demeter in pure Java) and
\urlh{http://sourceforge.net/projects/daj/}{DAJ} (Demeter for AspectJ)
from MS student projects.

\item[1992--1995]
{\large Software developer, Pure Software}, Sunnyvale, CA \\
Member of the development team for
\urlh{http://www.ibm.com/software/awdtools/purify/unix/}{Purify}, a
memory corruption and leak detection tool for C/C++ programs in
Solaris.  Worked on the reporting back-end, GUI, and common library
code for Purify, PureLink, and PureCoverage products.
\end{description}

\section*{Education}

\begin{description}
\item[1995--2005] 
{\large Ph.D. Computer Science, Northeastern University}, Boston, MA \\
Designed and implemented the
\urlh{http://socrates-lang.sf.net/}{Socrates} programming language
(embedded in PLT Scheme), which unifies and generalizes
object-oriented and aspect-oriented programming language mechanisms
for separation of concerns using predicate dispatching and open
classes.

\item[1988--1992]
{\large B.A. Computer Science, University of California}, Berkeley, CA \\
Member of the
\urlh{http://www.xcf.berkeley.edu/}{eXperimental Computing Facility (XCF)}.
Re-implemented the
\urlh{http://www.cs.berkeley.edu/~bh/}{Berkeley Logo}
interpreter as an
\urlh{http://mitpress.mit.edu/sicp/full-text/book/book-Z-H-34.html}{explicit-control evaluator}
rather than a
\urlh{http://mitpress.mit.edu/sicp/full-text/book/book-Z-H-26.html}{meta-circular evaluator}.
\end{description}

%% The ``Skills'' section seems kind of random.  Most of them appear in the
%% specific descriptions above; the others should maybe be tied to
%% specific project descriptions?  Also, some distinction could be
%% made between things I am very comfortable with and things I've used
%% a few times.

%% \section*{Skills}

%% \begin{description}
%% \item[Languages] PLT Scheme, Common Lisp, Java, AspectJ, ECMAScript,
%%   Python, C++/STL, MOO
%% \item[Development tools] Emacs, Subversion, GNUmake, Ant, tcsh, bash
%% \item[Operating systems] Ubuntu Linux, Solaris, Windows XP
%% \item[Technologies] DOM, Google App Engine, Jabber-RPC, Java Swing,
%%   JHCore, JUnit, MusicBrainz, OpenID, PostgreSQL, Prototype.js, RDF, SOAP,
%%   SVG, Volity, WS-BPEL, WSDL, XML Schemas, XMPP, XPath
%% \end{description}

\section*{Other Projects}

\begin{description}
\item[2011]
{\large \urlh{http://ifcomp.org/comp11/play.php?id=186}{Last Day of Summer}} \\
Designed, wrote, and implemented an interactive fiction game for the
\urlh{http://ifcomp.org/comp11/}{2011 IF Competition} using
\urlh{http://inform7.com/}{Inform 7}.  It placed 25th out of 38
entries---not bad for a first-timer competing against many experienced
IF authors!  It was part of a four-game metapuzzle that won the
\urlh{http://ifwiki.org/index.php/XYZZY_Awards_2011}{XYZZY Award} for
Best Individual Puzzle.
\end{description}

\begin{description}
\item[2009--2010]
{\large \urlh{http://chugchanga-poll.appspot.com/}{Chugchanga-L Poll}} \\
Designed and implemented an Ajax application in Python, hosted on
\urlh{http://appengine.google.com/}{Google App Engine}, to allow
members of Chugchanga-L (a music discussion mailing list) to vote for
their favorite music releases of the year.  It uses the
\urlh{http://musicbrainz.org}{MusicBrainz} XML web service to
canonicalize votes by matching them to releases in the MusicBrainz
database.
\end{description}

\begin{description}
\item[2006--2010]
{\large \urlh{http://volity.net/}{Volity Games Network}} \\
Helped implement the Volity platform for multi-player online board
games.  Implemented a Jabber-RPC library in Java, and helped develop
Gamut, the standalone SVG client Java application.  Implemented a
server (``parlor'') for a board game of my own design,
\urlh{http://www.icehousegames.org/wiki/index.php?title=Pylon}{Pylon},
in Python.  Took over maintenance and development of the Volity web
client.
\end{description}

\end{document}
